Optimistic Atomic Multicast in One Communication Delay

### 2011-07-22 trunk/textos/optpaxos/draft_dropbox.tex
revised proofs 2.0

Processes communicate using reliable multicast, defined by the primitives \rmcast{}$(s,m)$ and \rmdel{}$(m)$, where $s$ is a set of groups and $m$ is a message. Reliable multicast guarantees that (i)~for any process $p$ and any message $m$, $p$ \rmdel{}s $m$ at most once, and only if $p \in s$ and $m$ was previously r-mcast \emph{(uniform integrity)}; (ii)~if a correct process $p$ r-mcasts $m$, then eventually all correct processes $q \in s$ r-deliver $m$ \emph{(validity)}; (iii)~if $p$ r-delivers $m$, then eventually all correct processes $q \in s$ r-deliver $m$ \emph{(uniform agreement)}; and (iv)~if $p$ r-mcasts $m$ and then $m'$, then no $q$ r-delivers $m'$ without first delivering $m$ \emph{(fifo order)}.

@@

A more formal description of the protocol is given in Algorithm \ref{algorithm:deliveryminimal}. We consider that three primitives are given: $getTime()$, which returns the current value of the local wallclock; $Propose(k, val)$, which proposes a value $val$ for the consensus instance of id $k$ within its group; and also $Decide(k, val)$, which is called when the consensus instance of id $k$ finishes. $Decide(k, val)$ is called for all the processes of the group that initiated it, when they learn that the value $val$ has been agreed upon in instance of id $k$. For the sake of simplicity, we assume that, for consensus instances within the same group, the values are decided in the same order of the instances id's\footnote{This can be easily done by delaying the callback of $Decide(k, val)$ while there is some unfinished consensus instance of id $k': k' < k$ from the same group.}. Finally, we also use a FIFO reliable multicast primitive \rmcastarg{$m$}{$groupSet$}, which \rmdel{}s $m$ to all the processes in all the groups in $groupSet$ in one communication step, in FIFO order (e.g. the one described in \cite{ufrmcast1delta}).

Moreover, each process $p$ of group $G$ keeps some lists of messages:
\begin{itemize}
  %\item \textit{optPending}, which contains the messages waiting to be \mbox{\textit{\opt{}ed}} (optimistically delivered);
  \item \textit{groupMsgs}, containing the messages waiting to be proposed by $p$;
  \item \textit{barPending}, with the messages ready to be \cons{}ed, but which may be waiting for barriers from the groups in $sendersTo(G)$;
  \item \textit{decided}, which contains the messages that have already been proposed and decided within $group(p)$;
  \item \textit{delivered}, which contains the messages that have been \cons{}ed already.
\end{itemize}

@@

  When a process in $G$ \amcast{}s $m$, it is first \rmcast{} to the other processes of $G$. Then, when $m$ is \rmdel{}ed at $p$, $p$ checks whether the latest consensus instance $I_{k}$ in which it participated, or is trying to start, has already been decided -- if not, $p$ enqueues $m$ in a $groupMsgs$ queue as the next message being proposed by it in the consensus instance $I_{k+1}$, so other tasks can keep being executed. Then, once $I_{k}$ has been decided, $p$ may start a new instance. Before that, all messages in $groupMsgs$ that have been already decided are discarded from $groupMsgs$. The rest is proposed as a batch in $I_{k+1}$.

  Once $m$ is decided, it is not immediately delivered to the application. Instead, $p$ checks whether some message $m_{prv} : m_{prv}.ts \geq m.ts$ has been decided previously. If that is the case, the value of $m.ts$ is changed to a value greater than the timestamp of any other message previously decided within $G$. Then, $m$ is inserted into a $barPending$ list for later being delivered, which will happen once every group $G'$ in $sendersTo(G)$ has already sent a message $barrier(G',G) = t_b$, such that $t_b \geq m.ts$. This is done because there could be a message $m'$ yet to come from another group $G'$, such that $m'.ts < m.ts$.

@@

    In this case, when $m$ is decided, $p$ \rmcast{}s $m$ to every \mbox{$p' \in (m.dst \setminus \{G\})$}. When $m$ is \rmdel{}ed by each $p' \in G'$, $p'$ checks whether it has ever inserted $m$ in its own $barPending$ list. If not, $p'$ inserts $m$ into $barPending$ and adjusts $barrier(G, G')$ to $m.ts$. As the sending of these messages from a group to another is done via a FIFO reliable channel, once a message $m$ from another group $G$ is \rmdel{}ed, every message $m' : m'.ts < m.ts$ from $G$ has also already been \rmdel{}ed.

  \end{itemize}
%   
%   In this case, when $m$ is received by $p$, $p$ checks whether the latest consensus instance $I_{k}$ in which it participated, or is trying to start, has already been decided -- if not, $p$ enqueues $m$ in a $groupMsgs$ queue as the next message being proposed by it in the consensus instance $I_{k+1}$, so other tasks can keep being executed. Then, once $I_{k}$ has been decided, $p$ may start a new instance. Before that, all messages in $groupMsgs$ that have been already decided are discarded from $groupMsgs$. The rest is proposed as a batch in $I_{k+1}$. Once $m$ is decided, it is not immediately delivered to the application. Instead, $p$ checks whether some message $m_{prv} : m_{prv}.ts \geq m.ts$ has been decided previously. If that is the case, the value of $m.ts$ is changed to a value greater than the timestamp of any other message previously decided within $G$. Then, $m$ is inserted into a $barPending$ list for later being delivered, which will happen once every group $G'$ in $sendersTo(G)$ has already sent a message $barrier(G',G) = t$, such that $t > m.ts$. This is done because there could be a message $m'$ yet to come from another group $G'$, such that $m'.ts < m.ts$.
% 
%   \item When $m$ is originated in $G$, but it has at least one group other than $G$ as a destination:
%   
%   In this case, when $m$ is received, $p$ tries to initiate a consensus instance within $G$ to decide $m$. If $p$ cannot start the proposal now, $m$ is enqueued in $groupMsgs$ for being proposed later along with other pending messages. Then, once $p$ may start a new instance, all messages in $groupMsgs$ that have been already decided are discarded from $groupMsgs$. The rest is proposed as a batch in the new consensus instance. Once any message $m$ is decided, $p$ checks whether some message $m_{prv} : m_{prv}.ts > m.ts$ has been decided previously. If that is the case, the value of $m.ts$ is changed to a value greater than the timestamp of any other message previously decided within $G$. Then, if $G \in m.dst$, $m$ is inserted in the $barPending$ list. Besides, when $m$ is decided, $p$ sends $m$ to every \mbox{$p' \in (m.dst \setminus \{G\})$}. When $m$ is received by each $p' \in G'$, $p'$ checks whether it has ever inserted $m$ in its own $barPending$ list. If not, $p'$ inserts $m$ into $barPending$ and adjusts $barrier(G, G')$ to $m.ts$. To ensure that, once a message $m$ is received from another group $G$, every message $m' : m'.ts<m.ts$ also from $G$ has already been received, every message is sent through a lossless FIFO channel\footnote{An ordinary TCP connection would be enough to provide such FIFO lossless channel. Here, we use FIFO reliable multicast.}.

@@

\blankline
\INDENT{Initialization}
  \STATE $k \leftarrow 0$, $nextProp \leftarrow 0$, $decided \leftarrow \emptyset$, $delivered \leftarrow \emptyset$, $groupMsgs \leftarrow \emptyset$, $barPending \leftarrow \emptyset$
  \INDENT{\textbf{for all} $h \in sendersTo(g)$ \textbf{do}}
    \STATE $barrier(h) \leftarrow (0,0)$ 
  \ENDINDENT
\ENDINDENT

\blankline
\INDENT{\textit{To} \amcast{} \textit{a message $m$}}
  \STATE $m.ts \leftarrow (getTime(),0)$
  \STATE \rmcastarg{$\{g\}$}{$m$} \label{algline:rmlocal}
\ENDINDENT

\blankline
\WHEN{\rmdelarg{$m$}}
  \IF{$g = m.src$}
    \STATE $groupMsgs \leftarrow groupMsgs \cup \{m\}$ \label{algline:addtogroupMsgs}
  \ELSIF{$m \notin barPending \wedge m \notin delivered$}
    \STATE $barPending \leftarrow barPending \cup \{m\}$ \label{algline:insbp1}
    \STATE $barrier(m.src) \leftarrow m.ts$ \label{algline:incbar}
  \ENDIF
\ENDWHEN

\blankline
\WHEN{$groupMsgs \setminus decided \neq \emptyset$}
  \STATE $k \leftarrow k + 1$
	\STATE $undecided \leftarrow prop \setminus decided$
	\STATE propose($k, undecided$) \label{algline:propose}
	\STATE \textbf{wait until} decide($k, msgSet$)
  \STATE $msgSet \leftarrow msgSet \setminus decided$
	\WHILE{$msgSet \setminus decided \neq \emptyset$}
		\STATE let $m$ be the message in $msgSet \setminus decided$ with smallest timestamp
		\STATE let $m'$ be the message in $decided$ with greatest timestamp
		\IF{$m'.ts > m.ts$}
			\STATE $m.ts \leftarrow (m'.ts.rtc, m'.ts.seq + 1)$
		\ENDIF     
		\IF{$g \in m.dst$} \label{algline:checkcons}
			\STATE $barPending \leftarrow barPending \cup \{m\}$  \label{algline:insbp2}
		\ENDIF
		\STATE $decided \leftarrow decided \cup \{m\}$  \label{algline:addtodecided}
		\STATE \rmcastarg{$m.dst \setminus \{g\}$}{$m$} \label{algline:rmothers}
	\ENDWHILE

@@

\blankline
%\WHEN{$\exists m \in barPending : \forall G' \in sendersTo(G): m.ts < barrier(G',G)$\\ $\wedge\ \nexists m' \in barPending : m'.ts < m.ts$} \label{algline:inbarPending}
\WHEN{$barPending \setminus delivered \neq \emptyset$} \label{algline:inbarPending}
	\STATE let $m$ be the message in $barPending \setminus delivered$ with smallest timestamp
	\IF{$\forall h \in sendersTo(g): m.ts < barrier(h)$}
%		\STATE $barPending \leftarrow barPending \setminus \{m\}$ \label{algline:rmfrombp}
		\IF {$m \neq null$}
		  \STATE \amdelarg{$m$} \label{algline:consdeliver}
		\ENDIF
		\STATE $delivered \leftarrow delivered \cup \{m\}$ \label{algline:addtodelivered}
	\ENDIF
\ENDWHEN
\blankline

@@

\blankline
\INDENT{\textit{When $\exists h: g \in sendersTo(h) \wedge getTime() - lastBarrierCreated(h) > barrierThreshold$}}
  \STATE $null \leftarrow$ empty message
  \STATE $null.ts \leftarrow (lastBarrierCreated(h) + barrierThreshold, 0)$ \label{algline:samenullid}
  \STATE $null.src \leftarrow g$
  \STATE $null.dst \leftarrow \{h\}$ 
  \STATE $groupMsgs \leftarrow groupMsgs \cup \{null\}$
  \COMMENT{saving that nothing was sent until null.ts}
\ENDINDENT

@@

It is necessary to guarantee that a $null$ message will eventually be proposed, decided, and a barrier will be sent to some group which might be needing it, so that progression is guaranteed. Therefore, this kind of messages are created by every process in the group, since if any one of them does not have it, such message might never be decided. %Besides, they are excluded from the $groupMsgs$ list only when decided in the group (l. \ref{algline:nullstays} of Algorithm \ref{algorithm:deliveryminimal}). Since they are obviously never delivered to the application, they can remain in such list and be decided even after some message with higher timestamp is decided. Not removing them unless they are decided ensures that they will be sent and that other groups will be able to increase their barrier values. % On the other hand, this brings another problem: if every process in $G$ creates a different $null$ message, many $null$ messages might have to be decided before a meaningful message.
To prevent the multiple $null$ messages -- created by different processes within the group -- of being decided, they could be created in a way such that the different processes can somehow figure out that two different $null$ messages are equivalent\footnote{This could be done by assuming no timestamp collisions and by using them to uniquely identify messages. Then, only one of the messages created with the same timestamp (l. \ref{algline:samenullid} of Algorithm \ref{algorithm:nullperiodic}) would be decided.}.

@@

When the process $p'$ of some group $G'$ receives a barrier request for a message $m$, %such that $G' \in blockers(m)$,
$p'$ knows that there are other groups depending on the barrier of $G'$ to deliver $m$. For that reason, it will immediately create a $null$ message with a timestamp equal to $m.ts$ and with $null.dst = m.dst$ $\cap$ $receiversFrom(G')$. Then, $p'$ will insert such message in its %$optPending$
$groupMsgs$ queue. % -- to ensure that, once it goes to $groupMsgs$, any message \mbox{$m': m'.ts < null.ts \wedge m'.src = G'$} is already there. As soon as $null.ts > now - w(p')$, the 
Once the $null$ message is proposed and decided in $G'$, each process of $G'$ will send it to every process in each group \mbox{$G \in m.dst \cap receiversFrom(G')$}. This way, any group which was waiting for a barrier from $G'$ to deliver $m$ will be able to do so as soon as it receives such $null$ message. The new delivery algorithm would be as described in \mbox{Algorithm {\ref{algorithm:nullondemand}}}.

@@

\blankline
\INDENT {Initialization}
  \STATE $k \leftarrow 0$, $nextProp \leftarrow 0$, $decided \leftarrow \emptyset$, $delivered \leftarrow \emptyset$, $groupMsgs \leftarrow \emptyset$, $barPending \leftarrow \emptyset$
  \INDENT{\textbf{for all} $G' \in sendersTo(G)$ \textbf{do}}
    \STATE $barrier(G',G) \leftarrow -\infty$ 
  \ENDINDENT
\ENDINDENT

@@

\blankline
\INDENT{\textit{When} \rmdelarg{$m'$}}
  \IF{$G = m'.src$}
    \STATE $groupMsgs \leftarrow groupMsgs \cup \{m'\}$
  \ELSIF{$m' \notin barPending \wedge m' \notin delivered$}
    \STATE $barrier(m'.src,G) \leftarrow m'.ts$ \label{algline:incbar}
    \IF {$G \in m'.dst$}
      \STATE $barPending \leftarrow barPending \cup \{m'\}$
    \ENDIF
    \IF{$G \in blockers(m') \wedge m' \neq null $}
      \STATE $null \leftarrow$ empty message
      \STATE $null.src \leftarrow G$
      \STATE $null.ts \leftarrow m'.ts$ \label{algline:nulltsmts}
      \STATE $null.dst \leftarrow m'.dst \cap receiversFrom(G)$
      \STATE $groupMsgs \leftarrow groupMsgs \cup \{null\}$
    \ENDIF
  \ENDIF
\ENDINDENT

\blankline
\INDENT{\textit{When $\exists m \in groupMsgs \wedge nextProp = k$}}
%  \STATE $groupMsgs \leftarrow groupMsgs \setminus \{m\}$
%  \IF {$m \notin decided \wedge \nexists m' \in decided: m'.ts > m.ts$}
    \STATE $groupMsgs \leftarrow groupMsgs \setminus decided$%(decided \cup \{m' : \exists m'' \in decided \wedge m'.ts < m''.ts \wedge m' \neq null\})$ 
    \label{algline:nullstays}
    \IF {$groupMsgs \neq \emptyset$}
      \STATE $nextProp \leftarrow k + 1$
      \STATE Propose($k$,$groupMsgs$)
    \ENDIF
%  \ENDIF
\ENDINDENT

\blankline
\INDENT{\textit{When} Decide($k'$,$msgSet$)}
%  \STATE $m.gs \leftarrow k$
  \INDENT{\textbf{while} $\exists m \in msgSet : (\forall m' \in msgSet : m \neq m' \Rightarrow m.ts < m'.ts)$ \textbf{do}}
    \STATE $msgSet \leftarrow msgSet \setminus \{m\}$
    \COMMENT{the messages are handled in ascending order of timestamp}
    \IF{$\exists m' \in decided : m'.ts \geq m.ts \wedge (\nexists m'' \in decided : m''.ts > m'.ts)$}
      \STATE $m.ts \leftarrow (m'.ts.rtc, m'.ts.seq + 1)$ \label{algline:chgts}
    \ENDIF     
    \IF{$G \in m.dst$} \label{algline:checkcons}
      \STATE $barPending \leftarrow barPending \cup \{m\}$
    \ENDIF
    \STATE $decided \leftarrow decided \cup \{m\}$
    \IF{$m \neq null$}
      \STATE \rmcastarg{$m$}{$(m.dst \setminus \{G\}) \cup blockers(m)$} \label{algline:barreq} \COMMENT{the message is sent to blockers(m), as a barrier request}
    \ELSE
      \STATE \rmcastarg{$m$}{$m.dst$} \COMMENT{no need to unblock $null$, besides it would create infinite messages}
    \ENDIF
  \ENDINDENT
  \STATE $nextProp \leftarrow k' + 1$
  \STATE $k \leftarrow k' + 1$
\ENDINDENT

\blankline
\INDENT{\textit{When $\exists m \in barPending : \forall G' \in sendersTo(G): m.ts < barrier(G',G)$\\ $\wedge\ \nexists m' \in barPending : m'.ts < m.ts$ }}
  \STATE $barPending \leftarrow barPending \setminus \{m\}$
  \IF{$m \neq null$}
    \STATE \consdel{$m$}
  \ENDIF
  \STATE $delivered \leftarrow delivered \cup \{m\}$

@@

However, the assumption may not hold, so deliveries made based on it have to be confirmed. There would be then two deliveries for each message: an optimistic one, done within one communication step, and a conservative one, which may follow one of the algorithms described earlier. The difference is that, after a message $m$ was \rmdel{}ed at $p$, instead of proposing it, $p$ inserts it into an $optPending$ list, which is sorted in ascending order of the timestamps of the messages in it. Then, $p$ waits until its own wallclock has a value greater than $m.ts.rtc + w(p)$, after which $p$ both proposes $m$ and \opt{}s (optimistically delivers) it. If \emph{A2} holds, then all messages sent by $m.src$ and received after $m$ has been \opt{}ed will have a timestamp greater than $m.ts$.

@@

%     \STATE $groupMsgs \leftarrow groupMsgs \cup \{m\}$

>>>>

Processes communicate using fifo reliable multicast, defined by the primitives \rmcast{}$(s,m)$ and \rmdel{}$(m)$, where $s$ is a set of groups and $m$ is a message. Reliable multicast guarantees that (i)~for any process $p$ and any message $m$, $p$ \rmdel{}s $m$ at most once, and only if $p \in s$ and $m$ was previously \rmcast{} \emph{(uniform integrity)}; (ii)~if a correct process $p$ \rmcast{}s $m$, then eventually all correct processes $q \in s$ \rmdel{} $m$ \emph{(validity)}; (iii)~if $p$ \rmdel{}s $m$, then eventually all correct processes $q \in s$ \rmdel{}s $m$ \emph{(uniform agreement)}; and (iv)~if $p$ \rmcast{}s $m$ and then $m'$, then no $q$ \rmdel{}s $m'$ without first \rmdel{}ing $m$ \emph{(fifo order)}.

@@

A more formal description of the protocol is given in Algorithm \ref{algorithm:deliveryminimal}. We consider that three primitives are given: $getTime()$, which returns the current value of the local wallclock; $propose_g(k, v)$, which proposes a value $v$ for the consensus instance of id $k$ within its group; and also $decide_g(k, v)$, which is called when the consensus instance of id $k$ finishes. $decide_g(k, v)$ is called for all the processes of the group that initiated it, when they learn that the value $val$ has been agreed upon in instance of id $k$. For the sake of simplicity, we assume that, for consensus instances within the same group, the values are decided in the same order of the instances id's\footnote{This can be easily done by delaying the callback of $decide_g(k, v)$ while there is some unfinished consensus instance of id $k': k' < k$ from the same group.}. Finally, we also use a fifo reliable multicast primitive \rmcastarg{$m$}{$groupSet$}, which \rmdel{}s $m$ to all the processes in all the groups in $groupSet$ in one communication step, in fifo order (e.g. the one described in \cite{ufrmcast1delta}).

Moreover, each process $p$ of group $g$ keeps some sets of messages:
\begin{itemize}
  %\item \textit{optPending}, which contains the messages waiting to be \mbox{\textit{\opt{}ed}} (optimistically delivered);
  \item \textit{\localmsgs}, containing the %local messages, i.e., the 
  messages that have been \amcast{} by some process of group $g$;
  \item \textit{\decided}, which contains the messages that have already been proposed and decided within $group(p)$;
  \item \textit{\stamped}, containing the messages sent to $g$ by any group; a final timestamp has already been assigned to each of these messages, although, in order to be \amdel{}ed, some of them may still need barriers from the groups in $sendersTo(G)$;
  \item \textit{\delivered}, which contains the messages that have been already \amdel{}ed by $p$.
\end{itemize}

@@

  When a process in $G$ \amcast{}s $m$, it is first \rmcast{} to the other processes of $G$. Then, when $m$ is \rmdel{}ed at $p$, $p$ checks whether the latest consensus instance $I_{k}$ in which it participated, or is trying to start, has already been decided -- if not, $p$ enqueues $m$ in a $\localmsgs$ queue as the next message being proposed by it in the consensus instance $I_{k+1}$, so other tasks can keep being executed. Then, once $I_{k}$ has been decided, $p$ may start a new instance. Before that, all messages in $\localmsgs$ that have been already decided are discarded from $\localmsgs$. The rest is proposed as a batch in $I_{k+1}$.

  Once $m$ is decided, it is not immediately delivered to the application. Instead, $p$ checks whether some message $m_{prv} : m_{prv}.ts \geq m.ts$ has been decided previously. If that is the case, the value of $m.ts$ is changed to a value greater than the timestamp of any other message previously decided within $G$. Then, $m$ is inserted into a $\stamped$ set for later being delivered, which will happen once every group $G'$ in $sendersTo(G)$ has already sent a message $barrier(G',G) = t_b$, such that $t_b \geq m.ts$. This is done because there could be a message $m'$ yet to come from another group $G'$, such that $m'.ts < m.ts$.

@@

    In this case, when $m$ is decided, $p$ \rmcast{}s $m$ to every \mbox{$p' \in (m.dst \setminus \{G\})$}. When $m$ is \rmdel{}ed by each $p' \in G'$, $p'$ checks whether it has ever inserted $m$ in its own $\stamped$ set. If not, $p'$ inserts $m$ into $\stamped$ and adjusts $barrier(G, G')$ to $m.ts$. As the sending of these messages from a group to another is done via a fifo reliable channel, once a message $m$ from another group $G$ is \rmdel{}ed, every message $m' : m'.ts < m.ts$ from $G$ has also already been \rmdel{}ed.

  \end{itemize}
%   
%   In this case, when $m$ is received by $p$, $p$ checks whether the latest consensus instance $I_{k}$ in which it participated, or is trying to start, has already been decided -- if not, $p$ enqueues $m$ in a $\localmsgs$ queue as the next message being proposed by it in the consensus instance $I_{k+1}$, so other tasks can keep being executed. Then, once $I_{k}$ has been decided, $p$ may start a new instance. Before that, all messages in $\localmsgs$ that have been already decided are discarded from $\localmsgs$. The rest is proposed as a batch in $I_{k+1}$. Once $m$ is decided, it is not immediately delivered to the application. Instead, $p$ checks whether some message $m_{prv} : m_{prv}.ts \geq m.ts$ has been decided previously. If that is the case, the value of $m.ts$ is changed to a value greater than the timestamp of any other message previously decided within $G$. Then, $m$ is inserted into a $\stamped$ set for later being delivered, which will happen once every group $G'$ in $sendersTo(G)$ has already sent a message $barrier(G',G) = t$, such that $t > m.ts$. This is done because there could be a message $m'$ yet to come from another group $G'$, such that $m'.ts < m.ts$.
% 
%   \item When $m$ is originated in $G$, but it has at least one group other than $G$ as a destination:
%   
%   In this case, when $m$ is received, $p$ tries to initiate a consensus instance within $G$ to decide $m$. If $p$ cannot start the proposal now, $m$ is enqueued in $\localmsgs$ for being proposed later along with other pending messages. Then, once $p$ may start a new instance, all messages in $\localmsgs$ that have been already decided are discarded from $\localmsgs$. The rest is proposed as a batch in the new consensus instance. Once any message $m$ is decided, $p$ checks whether some message $m_{prv} : m_{prv}.ts > m.ts$ has been decided previously. If that is the case, the value of $m.ts$ is changed to a value greater than the timestamp of any other message previously decided within $G$. Then, if $G \in m.dst$, $m$ is inserted in the $\stamped$ set. Besides, when $m$ is decided, $p$ sends $m$ to every \mbox{$p' \in (m.dst \setminus \{G\})$}. When $m$ is received by each $p' \in G'$, $p'$ checks whether it has ever inserted $m$ in its own $\stamped$ set. If not, $p'$ inserts $m$ into $\stamped$ and adjusts $barrier(G, G')$ to $m.ts$. To ensure that, once a message $m$ is received from another group $G$, every message $m' : m'.ts<m.ts$ also from $G$ has already been received, every message is sent through a lossless fifo channel\footnote{An ordinary TCP connection would be enough to provide such fifo lossless channel. Here, we use fifo reliable multicast.}.

@@

\blankline
\INDENT{Initialization}
  \STATE $k \leftarrow 0$, $\localmsgs \leftarrow \emptyset$, $\decided \leftarrow \emptyset$, $\stamped \leftarrow \emptyset$, $\delivered \leftarrow \emptyset$
  \INDENT{\textbf{for all} $h \in sendersTo(g)$ \textbf{do}}
    \STATE $barrier(h) \leftarrow (0,0)$ 
  \ENDINDENT
\ENDINDENT

\blankline
\INDENT{\textit{To} \amcast{} \textit{a message $m$}}
  \STATE $m.ts \leftarrow (getTime(),0)$ \label{algline:gettime}
  \STATE \rmcastarg{$\{g\}$}{$m$} \label{algline:rmlocal}
\ENDINDENT

\blankline
\WHEN{\rmdelarg{$m$}}
  \IF{$g = m.src$}
    \STATE $\localmsgs \leftarrow \localmsgs \cup \{m\}$ \label{algline:addtogroupmsgs}
  \ELSIF{$m \notin \stamped$} \label{algline:notbptwice}
    \STATE $\stamped \leftarrow \stamped \cup \{m\}$ \label{algline:insbp1}
    \STATE $barrier(m.src) \leftarrow m.ts$ \label{algline:incbar}
  \ENDIF
\ENDWHEN

\blankline
\WHEN{$\localmsgs \setminus \decided \neq \emptyset$}
  \STATE $k \leftarrow k + 1$
	\STATE $undecided \leftarrow \localmsgs \setminus \decided$ \label{algline:undecided}
	\STATE propose$_g$($k, undecided$) \label{algline:propose}
	\STATE \textbf{wait until} decide$_g$($k, msgSet$) \label{algline:decide}
	\WHILE{$msgSet \setminus \decided \neq \emptyset$}
		\STATE let $m$ be the message in $msgSet \setminus \decided$ with smallest timestamp \label{algline:decidedloopbegin}
		\STATE let $m'$ be the message in $\decided$ with greatest timestamp
		\IF{$\exists m' \wedge m'.ts > m.ts$}
			\STATE $m.ts \leftarrow (m'.ts.rtc, m'.ts.seq + 1)$ \label{algline:changetsafterdecision}
		\ENDIF     
		\IF{$g \in m.dst$} \label{algline:checkcons}
			\STATE $\stamped \leftarrow \stamped \cup \{m\}$  \label{algline:insbp2}
		\ENDIF
		\STATE $\decided \leftarrow \decided \cup \{m\}$  \label{algline:addtodecided}
		\STATE \rmcastarg{$m.dst \setminus \{g\}$}{$m$} \label{algline:rmothers}
	\ENDWHILE

@@

\blankline
%\WHEN{$\exists m \in \stamped : \forall G' \in sendersTo(G): m.ts < barrier(G',G)$\\ $\wedge\ \nexists m' \in \stamped : m'.ts < m.ts$} \label{algline:in\stamped}
\WHEN{$\stamped \setminus \delivered \neq \emptyset$} \label{algline:inammulticast}
	\STATE let $m$ be the message in $\stamped \setminus \delivered$ with smallest timestamp \label{algline:condel1}
	\IF{$\forall h \in sendersTo(g): m.ts < barrier(h)$} \label{algline:condel2}
%		\STATE $\stamped \leftarrow \stamped \setminus \{m\}$ \label{algline:rmfrombp}
		  \STATE \amdelarg{$m$} \label{algline:consdeliver}
		\STATE $\delivered \leftarrow \delivered \cup \{m\}$ \label{algline:addtodelivered}
	\ENDIF
\ENDWHEN
\blankline

@@

\blankline
\INDENT{\textit{When $\exists h: g \in sendersTo(h) \wedge getTime() - lastBarrierCreated(h) > barrierThreshold$}}
  \STATE $null \leftarrow$ empty message
  \STATE $null.ts \leftarrow (lastBarrierCreated(h) + barrierThreshold, 0)$ \label{algline:samenullid}
  \STATE $null.src \leftarrow g$
  \STATE $null.dst \leftarrow \{h\}$ 
  \STATE $\localmsgs \leftarrow \localmsgs \cup \{null\}$ \label{algline:enqnull}
  \COMMENT{saving that nothing was sent until null.ts}
\ENDINDENT

@@

It is necessary to guarantee that a $null$ message will eventually be proposed, decided, and a barrier will be sent to some group which might be needing it, so that progression is guaranteed. Therefore, this kind of messages are created by every process in the group, since if any one of them does not have it, such message might never be decided. %Besides, they are excluded from the $\localmsgs$ set only when decided in the group (l. \ref{algline:nullstays} of Algorithm \ref{algorithm:deliveryminimal}). Since they are obviously never delivered to the application, they can remain in such set and be decided even after some message with higher timestamp is decided. Not removing them unless they are decided ensures that they will be sent and that other groups will be able to increase their barrier values. % On the other hand, this brings another problem: if every process in $G$ creates a different $null$ message, many $null$ messages might have to be decided before a meaningful message.
To prevent the multiple $null$ messages -- created by different processes within the group -- of being decided, they could be created in a way such that the different processes can somehow figure out that two different $null$ messages are equivalent\footnote{This could be done by assuming no timestamp collisions and by using them to uniquely identify messages. Then, only one of the messages created with the same timestamp (l. \ref{algline:samenullid} of Algorithm \ref{algorithm:nullperiodic}) would be decided.}.

@@

When the process $p'$ of some group $G'$ receives a barrier request for a message $m$, %such that $G' \in blockers(m)$,
$p'$ knows that there are other groups depending on the barrier of $G'$ to deliver $m$. For that reason, it will immediately create a $null$ message with a timestamp equal to $m.ts$ and with $null.dst = m.dst$ $\cap$ $receiversFrom(G')$. Then, $p'$ will insert such message in its %$optPending$
$\localmsgs$ queue. % -- to ensure that, once it goes to $\localmsgs$, any message \mbox{$m': m'.ts < null.ts \wedge m'.src = G'$} is already there. As soon as $null.ts > now - w(p')$, the 
Once the $null$ message is proposed and decided in $G'$, each process of $G'$ will send it to every process in each group \mbox{$G \in m.dst \cap receiversFrom(G')$}. This way, any group which was waiting for a barrier from $G'$ to deliver $m$ will be able to do so as soon as it receives such $null$ message. The new delivery algorithm would be as described in \mbox{Algorithm {\ref{algorithm:nullondemand}}}.

@@

\blankline
\INDENT {Initialization}
  \STATE $k \leftarrow 0$, $nextProp \leftarrow 0$, $\decided \leftarrow \emptyset$, $\delivered \leftarrow \emptyset$, $\localmsgs \leftarrow \emptyset$, $\stamped \leftarrow \emptyset$
  \INDENT{\textbf{for all} $G' \in sendersTo(G)$ \textbf{do}}
    \STATE $barrier(G',G) \leftarrow -\infty$ 
  \ENDINDENT
\ENDINDENT

@@

\blankline
\INDENT{\textit{When} \rmdelarg{$m'$}}
  \IF{$G = m'.src$}
    \STATE $\localmsgs \leftarrow \localmsgs \cup \{m'\}$
  \ELSIF{$m' \notin \stamped \wedge m' \notin \delivered$}
    \STATE $barrier(m'.src,G) \leftarrow m'.ts$ \label{algline:incbar}
    \IF {$G \in m'.dst$}
      \STATE $\stamped \leftarrow \stamped \cup \{m'\}$
    \ENDIF
    \IF{$G \in blockers(m') \wedge m' \neq null $}
      \STATE $null \leftarrow$ empty message
      \STATE $null.src \leftarrow G$
      \STATE $null.ts \leftarrow m'.ts$ \label{algline:nulltsmts}
      \STATE $null.dst \leftarrow m'.dst \cap receiversFrom(G)$
      \STATE $local\_msgs \leftarrow local\_msgs \cup \{null\}$
    \ENDIF
  \ENDIF
\ENDINDENT

\blankline
\INDENT{\textit{When $\exists m \in \localmsgs \wedge nextProp = k$}}
%  \STATE $\localmsgs \leftarrow \localmsgs \setminus \{m\}$
%  \IF {$m \notin \decided \wedge \nexists m' \in \decided: m'.ts > m.ts$}
    \STATE $\localmsgs \leftarrow \localmsgs \setminus \decided$%(\decided \cup \{m' : \exists m'' \in \decided \wedge m'.ts < m''.ts \wedge m' \neq null\})$ 
    \label{algline:nullstays}
    \IF {$\localmsgs \neq \emptyset$}
      \STATE $nextProp \leftarrow k + 1$
      \STATE propose$_g$($k$,$\localmsgs$)
    \ENDIF
%  \ENDIF
\ENDINDENT

\blankline
\INDENT{\textit{When} decide$_g$($k'$,$msgSet$)}
%  \STATE $m.gs \leftarrow k$
  \INDENT{\textbf{while} $\exists m \in msgSet : (\forall m' \in msgSet : m \neq m' \Rightarrow m.ts < m'.ts)$ \textbf{do}}
    \STATE $msgSet \leftarrow msgSet \setminus \{m\}$
    \COMMENT{the messages are handled in ascending order of timestamp}
    \IF{$\exists m' \in \decided : m'.ts \geq m.ts \wedge (\nexists m'' \in \decided : m''.ts > m'.ts)$}
      \STATE $m.ts \leftarrow (m'.ts.rtc, m'.ts.seq + 1)$ \label{algline:chgts}
    \ENDIF     
    \IF{$G \in m.dst$} \label{algline:checkcons}
      \STATE $\stamped \leftarrow \stamped \cup \{m\}$
    \ENDIF
    \STATE $\decided \leftarrow \decided \cup \{m\}$
    \IF{$m \neq null$}
      \STATE \rmcastarg{$m$}{$(m.dst \setminus \{G\}) \cup blockers(m)$} \label{algline:barreq} \COMMENT{the message is sent to blockers(m), as a barrier request}
    \ELSE
      \STATE \rmcastarg{$m$}{$m.dst$} \COMMENT{no need to unblock $null$, besides it would create infinite messages}
    \ENDIF
  \ENDINDENT
  \STATE $nextProp \leftarrow k' + 1$
  \STATE $k \leftarrow k' + 1$
\ENDINDENT

\blankline
\INDENT{\textit{When $\exists m \in \stamped : \forall G' \in sendersTo(G): m.ts < barrier(G',G)$\\ $\wedge\ \nexists m' \in \stamped : m'.ts < m.ts$ }}
  \STATE $\stamped \leftarrow \stamped \setminus \{m\}$
  \IF{$m \neq null$}
    \STATE \consdel{$m$}
  \ENDIF
  \STATE $\delivered \leftarrow \delivered \cup \{m\}$

@@

However, the assumption may not hold, so deliveries made based on it have to be confirmed. There would be then two deliveries for each message: an optimistic one, done within one communication step, and a conservative one, which may follow one of the algorithms described earlier. The difference is that, after a message $m$ was \rmdel{}ed at $p$, instead of proposing it, $p$ inserts it into an $optPending$ set, which is sorted in ascending order of the timestamps of the messages in it. Then, $p$ waits until its own wallclock has a value greater than $m.ts.rtc + w(p)$, after which $p$ both proposes $m$ and \opt{}s (optimistically delivers) it. If \emph{A2} holds, then all messages sent by $m.src$ and received after $m$ has been \opt{}ed will have a timestamp greater than $m.ts$.

@@

%     \STATE $\localmsgs \leftarrow \localmsgs \cup \{m\}$
